Assessing reliability of an online behavioural laterality battery: a pre-registered

### 2020-07-20 online_laterality_reliability.Rmd

Our final prediction related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult gather large amounts of data to establish test-retest reliability. For instance, Van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of measurement. In the preregistration, we predicted that for right-handers, a model (a) where the two language tasks are correlated with each other but not with the chimeric faces will provide better fit to the data than a model (b) in which laterality indices are correlated between the language and chimeric face tasks, or a base model (c) where each task is independently lateralised. For left-handers, we anticipated that that model (c) may be a better fit to the data, indicating that laterality is dissociated across the different tasks. However, we were not able to test this planned analysis with the current data set as one language task had low reliability. Thus, we confined our analysis to presentation of correlations between laterality indices from the four measures.

@@

Each trial began with a fixation cross presented centrally for 800 ms. Participants were instructed to keep their eyes on the central fixation point. Then a foveally presented word would appear for 200 ms. Word stimuli were presented in black lowercase Tahoma font on a white background. Font was scaled to 250% equating to 28pt font. The foveally presented word was replaced with bilaterally presented images at approximately 2.5 deg. of visual angle either side of fixation. The eccentricity of images was maintained using Gorilla’s scaling tool. The bilaterally presented images were then presented for 150 ms. At stimulus offset, participants indicated whether the image in the LVF or RVF rhymed with the foveally presented word. Participants were instructed to press ‘S’ to indicate that the foveally presented word rhyme with the image in the LVF and ‘K’ if it rhymed with the image in the RVF. If participants did not detect a rhyme, they were instructed to press the space bar. The response triggered the next trial.

@@

```{r read_dat, echo- FALSE, include= FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
# read in data from csv
LIs <- read.csv('https://osf.io/x5jd2/download',stringsAsFactors = F)
```

```{r prepdat,echo=FALSE,include=FALSE, comment=FALSE, warning=FALSE}
# remove those without day 2
# reformat NA
LIs$day2.complete <- ifelse(is.na(LIs$day2.complete), 0, LIs$day2.complete)
# remove
LIs_c <- LIs[LIs$day2.complete == 1,]
LIs <- LIs[LIs$day2.complete == 1,]
# lextale out df
LIs_no <- LIs[LIs$lexOut != "keep",]
# now remove them from LI
LIs <- LIs[LIs$lexOut == "keep",]
# table of counts for footed and sigh
short_dat <- dplyr::select(LIs, "subject", "handedness", "footedness", "miles", "porta")
long_dat <- tidyr::gather(short_dat, task, response, footedness:porta, factor_key=TRUE)
# make table
tab <- as.data.frame(table(long_dat$handedness, long_dat$task, long_dat$response))
tab <- tab %>% 
  rename(
    Handedness = Var1,
    Task = Var2,
    Response= Var3,
    Count= Freq
    )
split <- ggplot(data=tab, aes(x=Response, y=Count, fill=Handedness)) +
  geom_bar(stat="identity", position=position_dodge()) + facet_grid(.~Task) +
    geom_text(aes(label=Count), vjust=-0.3, position = position_dodge(0.9), size=3.5)+
  theme_classic() + theme(legend.position = "top") + ylab("Count") + xlab("") + ylim(0,220)
ggsave(
  "handed_sight.png",
  plot = split,
  width = 6, height = 4,
  dpi = 300
)
#Add a category of handedness that distinguishes extreme left from other left-handers
#In preregistration, extreme left defined as EHI laterality -90 or less.
LIs$handed3 <- LIs$handedness
w<-which(LIs$index_EHI<(-89))
LIs$handed3[w]<-'Extreme.left'
```
<!---Figure was not so easy to read : I think table preferable here--->
```{r simplifiedlatdescriptives}
#Miles and Porta are identical so only report one.
short_dat$cat3<-'R'
short_dat$cat3[short_dat$handedness=='Left']<-'L'
short_dat$temp<-'R'
short_dat$temp[short_dat$footedness=='Left']<-'L'
short_dat$temp[short_dat$footedness=="Don't know"]<-'X'
short_dat$cat3<-paste0(short_dat$cat3,short_dat$temp)
short_dat$temp<-'R'
short_dat$temp[short_dat$miles=='Left']<-'L'
short_dat$temp[short_dat$miles=="Don't know"]<-'X'
short_dat$cat3<-paste0(short_dat$cat3,short_dat$temp)
t<-table(short_dat$cat3)
lattab<-data.frame(matrix(NA,nrow=6,ncol=5))
colnames(lattab)<-c('Foot.Eye','R.handers.N','R.handers.%','L.handers.N','L.handers.%')
lattab$Foot.Eye<-c('RR','RL','LR','LL','Other','Total')
lattab$R.handers.N[6]<-sum(t[9:16])
lattab$R.handers.N[1:4]<-as.vector(t[c(13,12,10,9)])
lattab$R.handers.N[5]<-lattab$R.handers.N[6]-sum(lattab$R.handers.N[1:4])
lattab$L.handers.N[6]<-sum(t[1:8])
lattab$L.handers.N[1:4]<-as.vector(t[c(5,4,2,1)])
lattab$L.handers.N[5]<-lattab$L.handers.N[6]-sum(lattab$L.handers.N[1:4])
sumR<-lattab$R.handers.N[6]
sumL<-lattab$L.handers.N[6]
lattab$`R.handers.%`<-round(100*lattab$R.handers.N/sumR,1)
lattab$`L.handers.%`<-round(100*lattab$L.handers.N/sumL,1)
lattab$`R.handers.%`[6]<-NA
lattab$`L.handers.%`[6]<-NA

@@

##### Data cleaning
For each task we adopted the following data cleaning procedures. Applying each of these procedures meant that the number of participants left in the final analyses could vary by task. First, for all tasks, we removed trials in which reaction times were shorter than 200 ms. Outlier reaction times were identified using Hoaglin and Iglewicz’s (1987) procedure, whereby outliers are defined as reaction times that are 1.65 times the difference between the first and third quantiles (1.65 * (Q3-Q1)) below or above the first and third quantile values (e.g: lower limit = Q1 – 1.65 x (Q3-Q1); upper limit = Q3 + 1.65 x (Q3-Q1)). Our differences criterion was set to 1.65 instead of the typical 2.2 because we expected to obtain asymmetric distributions for reaction time, since reaction time is always non-negative, and so most outliers would be in one tail of the distribution. The Hoaglin and Iglewicz procedure was applied to each participants’ data separately. After applying each data trimming procedure, participants left with less than 80% of trials were removed.

For the chimeric faces task, we removed participants if they scored lower than 75% on trials in which an identical emotion was shown in each hemiface, and for the dichotic listening task, we removed those who scored less than 75% correct on trials where the same sound was played to each ear.  These data trimming procedures left data for `r nrow(LIs)-nrow(LIs[LIs$exclude== 1 | LIs$exclude== 12,])` participants on the chimeric faces `r nrow(LIs)-nrow(LIs[LIs$exclude== 2 | LIs$exclude== 12,])` participants on the dichotic listening task.

For the rhyme decision task, we removed participants who scored less than 50% accuracy in either visual field. Furthermore, we removed participants who scored less than 75% accuracy on the catch trials (trials in which there was no rhyme). These trimming procedures left data for `r nrow(LIs[LIs$exRDT== 0,])` participants on this task. We considered recruiting further participants to make up the numbers, but this seemed uneconomic as we already had more than 300 cases for the other tasks, and we could tell by this point that the test-retest reliability of the rhyme detection task was unsatisfactory; With `r nrow(LIs[LIs$exRDT== 0,])` participants, the test-retest reliability was *r*= `r round(cor(LIs[LIs$exRDT== 0,]$Day1.RDT_RT_LI, LIs[LIs$exRDT== 0,]$Day2.RDT_RT_LI, use = "complete.obs"),2)`, which is smaller than the *r*= 0.65 that we pre-registered as our criterion for a reliable task. Accordingly, we stopped data collection at this point and analysed the `r nrow(LIs)` participants with data on other tasks.

@@

For the rhyme task, we pre-registered that we would calculate LIs based on response time (RT) data, as accuracy during piloting was at ceiling. We specified that we would calculate LIs in two ways: (1) in the traditional fashion (100 x (Right – Left)/(Right + Left)); (2) to correspond to a *t*-value. This was achieved by conducting a *t*-test for each participant where accurate RTs are the dependent variable and visual field is the independent variable. This latter measure is highly correlated with the conventional laterality index but has the advantage that it can be used to identify participants who show a statistically significant RT advantage for one side, i.e. where the *t*-value on a 2-sided test has *p*< .05. Based on considerations from pilot data, we adopted log-transformation of RTs before computing *t*-values.

For the dichotic listening and chimeric faces, we specified that LI calculated would be based on trials in which participants correctly identified a stimulus. The count of correct responses that corresponded to each side was used to generate an accuracy LI in two ways: (1) in the traditional fashion (100 x (Right – Left)/(Right + Left)); (2) as a *z*-statistic for each participant, using z = (pR-.5)/sqrt(pR x pL/n), where pR is the proportion of correct responses in the RVF, pL is proportion of L responses in the LVF, and n is total LVF and RVF responses. This was analogous to the individual t-scores generated for RTs in the rhyme decision task, in allowing us to identify individuals with a significant bias to one side. Finally, LIs for the finger tapping task were calculated in the same manner, but using the number of sequences generated by the left and right hand as opposed to the proportion of correct responses in each visual field.

For the reliability assessment, we checked LIs for normality using a Shapiro-Wilk test. As the LIs were non-normal, we used a series of Spearman’s correlations to examine the correlation of coefficients for session 1 versus session 2. For tasks where the test-retest reliability was above .75 with the existing test length, we reanalysed the data using just the first N trials, reducing N until reliability becomes unsatisfactory (below *r*= .65). This was of interest in indicating the most efficient way to administer these tasks in future, while retaining good reliability. However, the full data (including all trials) was used for all subsequent analyses.

@@

##### Atypicality

To address hypothesis 2, that extreme left-handers will be more likely to show atypical laterality than right-handers, we categorised participants as extreme left-handers if they scored less than -90 and right-handers if they scored more than 90 of the Edinburgh Handedness Inventory. Atypicality was first defined in participants whose *t*- or *z*-score was significantly lateralised in the opposite direction from usual. If there were less than 20 cases defined as atypical using this method, we adjusted the cut-off on that task until we acquired 20 atypical participants. Once typicality was defined, we first analysed atypicality on the Rhyme Decision and Chimeric Faces task. Participants were categorised into four groups: typical on both; atypical on Rhyme Decision only; atypical on Chimeric Faces only, and atypical on both. A chi-square test was then used to test for association with the number of people in each of the four groups and the handedness classification from the Edinburgh Handedness Inventory. This was then be repeated for the Dichotic Listening and the Chimeric Faces tasks. A Bonferroni correction was applied to the outcome of each of the two chi-squared tests.

@@

```{r descriptives, echo=FALSE,include=FALSE, comment=FALSE, warning=FALSE}
# here we present the raw count and RT scores for each task
# read CF data
for_figs <- read.csv("for_figs.csv", na.strings = "NA")

# get CF
CF.short.dat <- for_figs[for_figs$exclude== 0 | for_figs$exclude== 2,]
  CF.short.dat <-CF.short.dat[CF.short.dat$task== "CF",]
# read Dichotic data
DL.short.dat <- for_figs[for_figs$exclude== 0 | for_figs$exclude== 1,]
  DL.short.dat <- DL.short.dat[DL.short.dat$task== "DL",]
# read rhyme data
RD.short.dat <- for_figs[for_figs$exRDT== 0,]
  RD.short.dat <- RD.short.dat[RD.short.dat$task== "RD",]
# read finger data
finger.short.dat <- read.csv("./data/finger.csv", na.strings = "NA")
# plots
CF <- ggplot(CF.short.dat, aes(x = side, y = accurate)) +
        ggpirate::geom_pirate(aes(colour = side)) +
        theme_classic() + ylab("Count of correct responses") + xlab("") + facet_wrap(.~Session)+ 
        ggtitle("C") + theme(plot.title = element_text(hjust = 0.5))
DL <- ggplot(DL.short.dat, aes(x = side, y = accurate)) +
        ggpirate::geom_pirate(aes(colour = side)) +
        theme_classic() + ylab("Count of correct responses") + xlab("") + facet_wrap(.~Session)+ 
        ggtitle("B") + theme(plot.title = element_text(hjust = 0.5))
RD <- ggplot(RD.short.dat, aes(x = side, y = RT)) +
        ggpirate::geom_pirate(aes(colour = side)) +
        theme_classic() + ylab("Reaction time (ms)") + xlab("") + facet_wrap(.~Session) + 
        ggtitle("A") + theme(plot.title = element_text(hjust = 0.5))
FT <- ggplot(finger.short.dat, aes(x = side, y = sequences)) +
        ggpirate::geom_pirate(aes(colour = side)) +
        theme_classic() + ylab("Count of correct sequences") + xlab("") + facet_wrap(.~Session)+ 
        ggtitle("D") + theme(plot.title = element_text(hjust = 0.5))
all.plot <- suppressMessages(suppressWarnings(ggpubr::ggarrange(RD, DL, CF, FT)))
all.plot
# save
ggsave(
  "all.vf.png",
  plot = all.plot,
  width = 6, height = 6,
  dpi = 300
)
```

<!---These figures aren't so good for showing laterality because they don't show how the L and R scores relate in individuals. Will try an alternative approach.--->

@@

makerawplot <- function(file1,file2,mysession,myname,mylim){

tempL<-filter(file1,Session==mysession)
tempR<-filter(file2,Session==mysession)
meanL<-mean(tempL$tempcol,na.rm=T)
meanR<-mean(tempR$tempcol,na.rm=T)
tempLR<-as.data.frame(cbind(tempL$tempcol,tempR$tempcol))
colnames(tempLR)<-c('Left','Right')
mytitle<-paste0(myname,"\n Session",mysession)
myg <- ggplot(tempLR, aes(x=Left, y=Right)) + geom_point(size=.5) +
  xlim(mylim)+ylim(mylim)+
    labs(title=mytitle)+
    geom_abline(intercept = 0, slope = 1, color="red", 
                 linetype="dotted")+
  geom_hline(yintercept=meanR,linetype='dotted')+geom_vline(xintercept=meanL,linetype='dotted')+
    theme(plot.title = element_text(size = 12))
return(myg)
}
```
```{r dorawplotfunction,echo=FALSE,include=FALSE, comment=FALSE, warning=FALSE}

RD.short.dat$tempcol<-RD.short.dat$RT
file1 <-filter(RD.short.dat,side=='left')
file2 <-filter(RD.short.dat,side=='right')
myname<-'Rhyme judge RT'
mysession<-1
mylim<-c(500,1500)
myg1<-makerawplot(file1,file2,mysession,myname,mylim)
myg2<-makerawplot(file1,file2,2,myname,mylim)

DL.short.dat$tempcol<-DL.short.dat$accurate
file1 <-filter(DL.short.dat,side=='left')
file2 <-filter(DL.short.dat,side=='right')
myname<-'Dichotic accuracy'
mysession<-1
mylim<-c(0,100)
myg3<-makerawplot(file1,file2,mysession,myname,mylim)
myg4<-makerawplot(file1,file2,2,myname,mylim)

CF.short.dat$tempcol<-CF.short.dat$accurate
#there are 96 stimuli with different faces, so 48 is pt of equality if all trials included
file1 <-filter(CF.short.dat,side=='left')
file2 <-filter(CF.short.dat,side=='right')
myname<-'Chimeric faces accuracy'
mylim<-c(0,96)
mysession<-1
myg5<-makerawplot(file1,file2,mysession,myname,mylim)
myg6<-makerawplot(file1,file2,2,myname,mylim)

finger.short.dat$tempcol<-finger.short.dat$sequences
file1 <-filter(finger.short.dat,side=='left')
file2 <-filter(finger.short.dat,side=='right')
myname<-'Finger tap sequences'
mysession<-1
mylim<-c(0,175)
myg7<-makerawplot(file1,file2,mysession,myname,mylim)
myg8<-makerawplot(file1,file2,2,myname,mylim)

ggarrange(myg1,myg2,myg3,myg4,myg5,myg6,myg7,myg8, ncol = 2, nrow = 4)
ggsave(
  "all.raw.png",
  width = 5, height = 8,
  dpi = 300
)
```

### Laterality indices: means and test-retest reliability

Accuracy on the rhyme task was high  (*M*= `r round(mean(RD.short.dat$p.corr, na.rm= TRUE),1)`%, *SD*= `r round(sd(RD.short.dat$p.corr, na.rm= TRUE),2)`%), and only correct response times were used for computing a laterality index. Average reaction times on this task are shown in Figure 3 as vertical and horizontal lines in a scatterplot showing right vs left visual field, separately for each session. A sloping dotted line shows the point where left and right RTs are equal; if there is faster performance for the right visual field, as predicted, then points should cluster below this line. Similar scatterplots are shown for selection of left and right responses for dichotic listening and chimeric faces. Note that for these two tasks, either side can give a correct response; the negative correlation between the two sides reflects the fact that the number of selections on the left will be inversely related to the number of correct responses on the right. For these tasks it is evident by inspection that most responses for dichotic listening cluster above the dotted line, indicating a right ear advantage, whereas most responses on chimeric faces cluster below the dotted line. Finally, we show the number of correct sequences with each hand on the finger-tapping task. Here, as expected, inspection of the data suggests more sequences with the right than the left hand on average.

@@

### Atypicality

We defined 'atypical laterality' as laterality that is in the opposite direction to the majority of participant, and which has an absolute *z*/*t*-score value greater than 1.96. Based on our pre-registered analysis, we divided left-handers into two groups (1) those with a score of -90 or below on the Edinburgh Handedness Inventory (c.f. Mazoyer et al., 2014), (2) the remaining left handers. While our analyses include only right handers and extreme left handers, we include the remaining left-handers in Table 2 for completeness.

>>>>

Our final prediction related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult gather large amounts of data to establish test-retest reliability. For instance, Van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of measurement. In the preregistration, we predicted that for right-handers, a model (a) where the two language tasks are correlated with each other but not with the chimeric faces will provide better fit to the data than a model (b) in which laterality indices are correlated between the language and chimeric face tasks, or a base model (c) where each task is independently lateralised. For left-handers, we anticipated that that model (c) may be a better fit to the data, indicating that laterality is dissociated across the different tasks.

@@

Each trial began with a fixation cross presented centrally for 800 ms. Participants were instructed to keep their eyes on the central fixation point. Then a foveally presented word would appear for 200 ms. Word stimuli were presented in black lowercase Tahoma font on a white background. Font was scaled to 250% equating to 28pt font. The foveally presented word was replaced with bilaterally presented images at approximately 2.5 deg. of visual angle either side of fixation. The eccentricity of images was maintained using Gorilla’s scaling tool. The bilaterally presented images were then presented for 150 ms. At stimulus offset, participants indicated whether the image in the LVF or RVF rhymed with the foveally presented word. Participants were instructed to press ‘S’ to indicate that the foveally presented word rhyme with the image in the LVF and ‘K’ if it rhymed with the image in the RVF. If participants did not detect a rhyme, they were instructed to press the space bar. The response triggered the next trial. Accuracy and response time were recorded.

@@

```{r read_dat, echo- FALSE, include= FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpubr)
library(corrr)
library(flextable)
options(knitr.kable.NA = '') #should suppress NA in tables
# read in data from csv
#LIs <- read.csv('https://osf.io/x5jd2/download',stringsAsFactors = F)
#We will replace with read from OSF, but better stick with LIs to ensure final version is the one we use 
allsum <- read.csv('./data/allsum.csv',stringsAsFactors = F)

```

```{r prepdat,echo=FALSE,include=FALSE, comment=FALSE, warning=FALSE}
# remove those without day 2 to make LIs file - we will just use this
# reformat NA
allsum$day2.complete <- ifelse(is.na(allsum$day2.complete), 0, allsum$day2.complete)
LIs <- allsum[allsum$day2.complete == 1,]

# remove
LIs_c <- LIs[LIs$day2.complete == 1,]

# lextale out df
LIs_no <- LIs[LIs$lexOut != "keep",]
# now remove them from LI
LIs <- LIs[LIs$lexOut == "keep",]
# table of counts for footed and sigh
short_dat <- dplyr::select(LIs, "subject", "handedness", "footedness", "miles", "porta")
long_dat <- tidyr::gather(short_dat, task, response, footedness:porta, factor_key=TRUE)
# make table
tab <- as.data.frame(table(long_dat$handedness, long_dat$task, long_dat$response))
tab <- tab %>% 
  rename(
    Handedness = Var1,
    Task = Var2,
    Response= Var3,
    Count= Freq
    )
split <- ggplot(data=tab, aes(x=Response, y=Count, fill=Handedness)) +
  geom_bar(stat="identity", position=position_dodge()) + facet_grid(.~Task) +
    geom_text(aes(label=Count), vjust=-0.3, position = position_dodge(0.9), size=3.5)+
  theme_classic() + theme(legend.position = "top") + ylab("Count") + xlab("") + ylim(0,220)
ggsave(
  "handed_sight.png",
  plot = split,
  width = 6, height = 4,
  dpi = 300
)
#Add a category of handedness that distinguishes extreme left from other left-handers
#In preregistration, extreme left defined as EHI laterality -90 or less.
LIs$handed3 <- LIs$handedness
w<-which(LIs$index_EHI<(-89))
LIs$handed3[w]<-'Extreme.left'
```
<!---Figure was not so easy to read : I think table preferable here--->
```{r simplifiedlatdescriptives}
#Miles and Porta are identical so only report one.
short_dat$cat3<-'R'
short_dat$cat3[short_dat$handedness=='Left']<-'L'
short_dat$temp<-'R'
short_dat$temp[short_dat$footedness=='Left']<-'L'
short_dat$temp[short_dat$footedness=="Don't know"]<-'X'
short_dat$cat3<-paste0(short_dat$cat3,short_dat$temp)
short_dat$temp<-'R'
short_dat$temp[short_dat$miles=='Left']<-'L'
short_dat$temp[short_dat$miles=="Don't know"]<-'X'
short_dat$cat3<-paste0(short_dat$cat3,short_dat$temp)
t<-table(short_dat$cat3)
lattab<-data.frame(matrix(NA,nrow=6,ncol=5))
colnames(lattab)<-c('Foot.Eye','R.handers.N','R.handers.%','L.handers.N','L.handers.%')
lattab$Foot.Eye<-c('RR','RL','LR','LL','Other','Total')
lattab$R.handers.N[6]<-sum(t[9:16])
lattab$R.handers.N[1:4]<-as.vector(t[c(13,12,10,9)])
lattab$R.handers.N[5]<-lattab$R.handers.N[6]-sum(lattab$R.handers.N[1:4])
lattab$L.handers.N[6]<-sum(t[1:8])
lattab$L.handers.N[1:4]<-as.vector(t[c(5,4,2,1)])
lattab$L.handers.N[5]<-lattab$L.handers.N[6]-sum(lattab$L.handers.N[1:4])
sumR<-lattab$R.handers.N[6]
sumL<-lattab$L.handers.N[6]
lattab$`R.handers.%`<-round(100*lattab$R.handers.N/sumR,1)
lattab$`L.handers.%`<-round(100*lattab$L.handers.N/sumL,1)
lattab$`R.handers.%`[6]<-NA
lattab$`L.handers.%`[6]<-NA

@@

##### Data cleaning
For each task we adopted the following data cleaning procedures. Applying each of these procedures meant that the number of participants left in the final analyses could vary by task. First, for all tasks, we removed trials in which response times were shorter than 200 ms. Response times were log-transformed to reduce skew, as specified in our pre-registration. Outlier reaction times were identified using Hoaglin and Iglewicz’s (1987) procedure, whereby outliers are defined as response times that are 1.65 times the difference between the first and third quantiles (1.65 * (Q3-Q1)) below or above the first and third quantile values (e.g: lower limit = Q1 – 1.65 x (Q3-Q1); upper limit = Q3 + 1.65 x (Q3-Q1)). Our differences criterion was set to 1.65 instead of the typical 2.2 because we expected to obtain asymmetric distributions for response time, with most outliers in one tail of the distribution. The Hoaglin and Iglewicz procedure was applied to each participant's data separately. After applying each data trimming procedure, participants left with less than 80% of trials were removed.

For the chimeric faces task, we removed participants if they scored lower than 75% on trials in which an identical emotion was shown in each hemiface, and for the dichotic listening task, we removed those who scored less than 75% correct on trials where the same sound was played to each ear.  These data trimming procedures left data for `r nrow(LIs)-nrow(LIs[LIs$exclude== 1 | LIs$exclude== 12,])` participants on the chimeric faces `r nrow(LIs)-nrow(LIs[LIs$exclude== 2 | LIs$exclude== 12,])` participants on the dichotic listening task. An additional 7 cases had to be removed because of a technical error that meant they had less than 80 trials presented.

For the rhyme decision task, we removed participants who scored less than 50% accuracy in either visual field. Furthermore, we removed participants who scored less than 75% accuracy on the catch trials (trials in which there was no rhyme). These trimming procedures left data for `r nrow(LIs[LIs$exRDT== 0,])` participants on this task. We considered recruiting further participants to make up the numbers, but this seemed uneconomic as we already had more than 300 cases for the other tasks, and we could tell by this point that the test-retest reliability of the rhyme detection task was unsatisfactory; With `r nrow(LIs[LIs$exRDT== 0,])` participants, the test-retest reliability was *r*= `r round(cor(LIs[LIs$exRDT== 0,]$Day1.RDT_RT_LI, LIs[LIs$exRDT== 0,]$Day2.RDT_RT_LI, use = "complete.obs"),2)`, which is smaller than the *r*= 0.65 that we pre-registered as our criterion for a reliable task. We plan to evaluate modified versions of this task in future, but decided to stop data collection at this point and analysed the `r nrow(LIs)` participants with data on other tasks.

For the finger-tapping task we added a data-cleaning step that had not been pre-registered, to remove participants where there were more than 20 keypresses across left and right hands that did not correspond to the  intended sequence. This could reflect that the individual had forgotten the target sequence, or was not co-operating with instructions, but scrutiny of cases where this occurred suggested that it usually arose because of technical problems. For instance, some individuals had repeated recording of the same keypress at short intervals, suggestive of a sticky keyboard. The cutoff of 20 was arbitrarily selected, and was successful in excluding cases with abnormally low scores in one session.

@@

We preregistered that we would calculate LIs in two ways: (1) in the traditional fashion (100 x (Right – Left)/(Right + Left)); (2) a standardized LI that involved computing a z-score that represented, for each individual, a comparison between left- and right-sided responses. In fact, for RT data we specified a t-score, but t and z converge when sample size exceeds 30, and for simplicity we refer to these as LI_z scores. These LI_z scores are very highly correlated with the traditional LI measures (correlations typically exceed .95), but can be used to provide a common metric for RT and accuracy measures when identifying individuals who are significantly lateralised. For example, using a conventional 2-sided alpha of .05, we can categorise individuals with an absolute LI_z greater than 1.96 as significantly lateralised.

For the rhyme task, we pre-registered that we would calculate LIs based on log response time (RT) data, as accuracy during piloting was at ceiling. To obtain LI_z we ran a *t*-test for each participant where accurate log RTs were the dependent variable and visual field was the independent variable.

For the dichotic listening and chimeric faces, we specified that LI calculated would be based on trials in which participants correctly identified a stimulus. The count of correct responses that corresponded to each side was used to generate an accuracy LI in two ways: (1) in the traditional fashion (100 x (Right – Left)/(Right + Left)); (2) as a *z*-statistic for each participant, using z = (pR-.5)/sqrt(pR x pL/n), where pR is the proportion of correct responses in the RVF, pL is proportion of L responses in the LVF, and n is total LVF and RVF responses. This method can yield extreme z-scores (for instance, if 83/89 responses are to the right, then z = 15.92), so scores were censored to be in the range -5 to +5.  Finally, LIs for the finger tapping task were calculated in the same manner,  using the number of sequences generated by the left and right hand as opposed to the proportion of correct responses in each visual field.

@@

For the reliability assessment, we checked LIs for normality using a Shapiro-Wilk test. As most LIs were non-normal, we used a series of Spearman’s correlations to examine the correlation of coefficients for session 1 versus session 2. For tasks where the test-retest reliability was above .75 with the existing test length, we reanalysed the data using just the first N trials, reducing N until reliability becomes unsatisfactory (below *r*= .65). This was of interest in indicating the most efficient way to administer these tasks in future, while retaining good reliability. However, the full data (including all trials) was used for all subsequent analyses.

##### Atypical laterality

To address hypothesis 2, that extreme left-handers will be more likely to show atypical laterality than right-handers, we categorised participants as extreme left-handers if they scored less than -90 and right-handers if they scored more than 90 on the Edinburgh Handedness Inventory. Our preregistration specified that atypicality would first be defined according to whether the participant's LI_z was significantly lateralised (with alpha = .05) in the opposite direction from the majority. If there were less than 20 cases defined as atypical using this method, we planned to adjust the cut-off on that task until we acquired 20 atypical participants.

For Rhyme Decision and Chimeric Faces task, participants were categorised into four groups: typical on both; atypical on Rhyme Decision only; atypical on Chimeric Faces only, and atypical on both. A chi-square test was then used to test for association with the number of people in each of the four groups and the handedness classification from the Edinburgh Handedness Inventory. This was then be repeated for the Dichotic Listening and the Chimeric Faces tasks. A Bonferroni correction was applied to the outcome of each of the two chi-squared tests.

@@

@@

makerawplot <- function(tempLR,mysession,myname,mylim){
  colnames(tempLR)<-c('Left','Right','Handedness')
  tempLR<-tempLR[!is.na(tempLR$Handedness), ] #remove rows with missing data
  tempLR$Handedness<-as.factor(tempLR$Handedness)
  droplevels(tempLR$Handedness)
  levels(tempLR$Handedness)=c('Right','Left')
  #myalpha<-c(1,.75) #haven't been able to adjust cols or pch this way. alpha works but one is always v faint regardless of setting
  meanL<-mean(tempLR$Left,na.rm=T)
  meanR<-mean(tempLR$Right,na.rm=T)
  mytitle<-paste0(myname,"\n Session",mysession)
  myg <- ggplot(tempLR, aes(x=Left, y=Right,col=Handedness)) + geom_point(size=.5) +
    xlim(mylim)+ylim(mylim)+
    labs(title=mytitle)+
    geom_abline(intercept = 0, slope = 1, color="red", 
                linetype="dotted")+
    geom_hline(yintercept=meanR,linetype='dotted')+geom_vline(xintercept=meanL,linetype='dotted')+
    theme(legend.position = "none") +
    theme(plot.title = element_text(size = 12)
    )
  myg<-myg+ scale_color_manual(values=c("black","red"))
  return(myg)
}
```
```{r dorawplotfunction,echo=FALSE,include=FALSE, comment=FALSE, warning=FALSE}

myname<-'Rhyme judge RT'
tempLR<-select(LIs[LIs$exRDT==0,],RDTlog1.Lraw,RDTlog1.Rraw,handpch)
#convert back to raw time
tempLR[,1:2]<-exp(tempLR[,1:2])
session<-1
mylim<-c(500,1500)
myg1<-makerawplot(tempLR,session,myname,mylim)

tempLR<-select(LIs[LIs$exRDT==0,],RDTlog2.Lraw,RDTlog2.Rraw,handpch)
tempLR[,1:2]<-exp(tempLR[,1:2]) #convert from log back to original scale for plot
session<-2
myg2<-makerawplot(tempLR,session,myname,mylim)

myname<-'Dichotic accuracy'
tempLR<-select(LIs[LIs$exclude<2,],DL1.Lraw,DL1.Rraw,handpch)
session<-1
mylim<-c(0,100)
myg3<-makerawplot(tempLR,session,myname,mylim)
tempLR<-select(LIs[LIs$exclude<2,],DL2.Lraw,DL2.Rraw,handpch)
session<-2
myg4<-makerawplot(tempLR,session,myname,mylim)

myname<-'Chimeric faces accuracy'
#LIs exclude of 1 or 12 indicates excluded from CF; retain exclude 0,2
tempLR<-select(LIs[LIs$exclude %in% c(0,2),],CF1.Lraw,CF1.Rraw,handpch)
session<-1
mylim<-c(0,100)
myg5<-makerawplot(tempLR,session,myname,mylim)
tempLR<-select(LIs[LIs$exclude %in% c(0,2),],CF2.Lraw,CF2.Rraw,handpch)
session<-2
myg6<-makerawplot(tempLR,session,myname,mylim)

myname<-'N finger sequences'
tempLR<-select(LIs,finger.L1,finger.R1,handpch)
session<-1
mylim<-c(50,180)
myg7<-makerawplot(tempLR,session,myname,mylim)
tempLR<-select(LIs,finger.L2,finger.R2,handpch)
session<-2
myg8<-makerawplot(tempLR,session,myname,mylim)

ggarrange(myg1,myg2,myg3,myg4,myg5,myg6,myg7,myg8, ncol = 2, nrow = 4,common.legend=T,legend="bottom")
ggsave(
  "./plots/all.raw.png",
  width = 5, height = 8,
  dpi = 300
)
```

```{r computepcorrRDT}
#revised this, as p.corr values now in allsum
RDT.short.dat<-allsum[allsum$exRDT==0,]
tempRDT <- select(RDT.short.dat,RDT1.pcorrL,RDT1.pcorrR,RDT2.pcorrL,RDT2.pcorrR)
allmeanpcorr <- mean(colMeans(tempRDT,na.rm=T))
allsdpcorr <- sd(c(tempRDT$RDT1.pcorrL,tempRDT$RDT1.pcorrR,tempRDT$RDT2.pcorrL,tempRDT$RDT2.pcorrR),na.rm=T)
```
### Significance of lateralisation on each task

For a preliminary impression of the data, Figure 3 shows scatterplots indicating scores on left and right sides for each task, with handedness colour-coded. For each task, mean for each side is shown as vertical and horizontal dotted lines, and equal performance on the two sides corresponds to the sloping red dotted line: thus an impression of whether the task is lateralised can be obtained by considering whether points cluster around the sloping lines symmetrically.

For the rhyme task, accuracy was high (*M* = `r round(allmeanpcorr,1)`%, *SD* = `r round(allsdpcorr,1)` %),and only correct responses were used for computing mean log RTs (shown transformed back to original ms scale in the plot). For dichotic listening and chimeric faces a correct response can be given from either side; it follows that there will be a negative correlation between the two sides reflecting the fact that the number of selections on the left will be inversely related to the number of correct responses on the right. For these tasks it is evident by inspection that most responses for dichotic listening cluster above the dotted line, indicating a right ear advantage, whereas most responses on chimeric faces cluster below the dotted line, indicating a left hemi-face advantage. Finally, we show the number of correct sequences with each hand on the finger-tapping task. Here, a strong impact of handedness is evident on inspection, with left-handers generally more accurate with the left hand, and right-handers more accurate with the right hand.

@@

### Atypical laterality and handedness

According to Mazoyer et al (2014), extreme left-handers are more likely than other handedness groups to have reversed laterality compared to the rest of the population. Our preregistered analysis accordingly focused on comparing extreme left-handers (those with a score of -90 or below on the Edinburgh Handedness Inventory) with right-handers. We defined 'atypical laterality' as laterality that is in the opposite direction to the majority of participants, with an absolute LI_z value greater than 1.96. Although the focus of our planned analyses included only right handers and extreme left handers, we show data for  the remaining left-handers for completeness.
